Network Origins of Social Distinction

### 2025-07-26 manuscript.tex
Updates from Overleaf

\newpage
\begin{abstract}	
This paper introduces the concept of \textit{distinction centrality}, developing a network metric designed to capture it. The metric is theoretically inspired by Bourdieu's theory of agonistic competition for distinction in social fields, where other actors and their actions provide opportunities and constraints to maximize social positioning. The basic idea behind the network metric of distinction is to subtract the \textit{autonomy} of each actor within the network from their \textit{constraint}, such that highly autonomous (and therefore less constrained) actors have more distinction. Autonomy is given by the focal actor's Eigenvector Centrality score (recording connections to well-placed others), and constraint is defined by the average Eigenvector Centrality score of each actor's neighbors in a hypothetical network where the focal actor is removed (indicating how others would benefit from that actor's exiting the network). A scalar is then applied to create mathematical and sociological equivalence between autonomy and constraint. The resulting node-level measures create a measure of \textit{distinction}, where the uniqueness of an actor's relations is more heavily rewarded than under the standard Eigenvector Centrality score. We analyze two classic datasets, Padgett and Ansell's \citeyearpar{padgett1993robust} Medici Network and Zachary's \citeyearpar{zachary1977information} Karate Club, to demonstrate that the observed actions of the agents involved are consistent with the maximization of distinction. We close by outlining how the resulting distinction measure can be used to both forecast network transformations and anticipate competitive changes resulting from external network shocks.    
\end{abstract}

@@

Bourdieu \citeyearpar{bourdieu1984distinction} most famously used the conceptualization of distinction to explain the persistence of the relationship between class and cultural hierarchies. Cultural elites in France distinguished themselves from their inferiors by adopting differentiated cultural practices that had undergone a process of consecration, which set them apart from the common. There is a strong correlation between those in the position of cultural and economic elites. Still, such a correlation is not guaranteed at the individual level, even if it is relevant at the field level.

While largely a concept used by those who research cultural practices, the pursuit of distinction is a general expression of the interaction between status and uniqueness within a field of positions \citep{anheier1995forms}. For example, while cultural practices may be a marker of difference, the absence of association with others outside the milieu is also a matter of agency within the milieu, a question of the constraints imposed by the set of relationships within which the actor is embedded. For Bourdieu, the struggle for position in a field was a product of a field of action and a field of forces \citep{martin2003field}. The field of action refers to the agency of actors actively advancing their interests. In contrast, the field of forces describes the counteracting pressures faced by the other participants whose interests would be upset by positional transformations.

In relation to theory, network status is another name for \textit{influence} \citep{friedkin1991theoretical}, and uniqueness for \textit{autonomy} in a social environment \citep{burt1980autonomy}. Translated to network terms, Bourdieu's field of action can be understood as the status hierarchy of the network, while the field of struggle is reflective of the constraints neighbors put on an actor. An actor with high influence is well-connected to others who can be mobilized for effective action. An actor with high autonomy is not constrained by their pattern connections or forced into action by them \citep{burt76}. We often use context-specific words to denote position in the hierarchy and ability to act within that hierarchy. Nevertheless, we use ``status" and ``autonomy" below as general terms to describe the ability of actors to take action versus the level of imposition from the forces of constraint.

In developing a broader understanding and measure of social distinction, actor-level incentives and structural impediments to action become apparent. When we understand the underlying motivation actors have to proactively reshape their network connections so as to achieve more distinct positions, historical patterns of network change can be better accounted for and sometimes even properly forecasted.

\subsection{Position as the Relational Basis for a Distinction Measure}
Bourdieu's separation of distinction into habitus and position is not a sociological problem at all, but rather creates the analytical problem of how an actor's network position generates dispositions and how these dispositions create opportunities for network mobility. Both dispositions and positions are fundamentally relational outcomes in sociological terms; however, when creating a network measure of distinction, only the relational basis of position can be considered.

@@

Which tells us that the distinction of the peripheral node in the star graph is proportional to its own (scale) status, discounted by the distinction of the star node divided by the same scaling factor.

@@

\input{Tabs/clique}

@@

\subsubsection{Circle Graphs and Triadic Closure}
%NOTES FROM 7|19|25 Meeting
%***Triadic closure as distinction maximization*** 
%Strogatz is random, preferential attachment of closure
%Status homophilous (enh) attachment vs preferential vs random 
%I wonder how much of balance theory is really just distinction maximization?
%Non psychological theory polarity

%Triadic closure 3 versions
\textit Triadic closure is an observed regularity in which a connection is likely to form between two actors who, although not originally connected, are first connected to a common alter (Simmel 1908). In essence, friends of friends have an increased likelihood of also becoming friends. The causes of this regularity, and the reasons for its occasional failure, are largely based on consideration of valence. Positive or negative affect drive the resulting triadic relations. The original consideration was for the cognitive balance of the individual, or the psychological congruity of affect by an individual for two objects that they see as causally connected. As such, the valenced relationship of any three actors are brought into combined consideration because any two are considered affectively combined through their joint combination as "the parts of a causal unit (Heider 1946:107)."

\citep{huang2018will}
\citep{yoon2013exchange}
\citep{mollenhorst2011shared} 
\citep{feld1981focused}
\citep{feld1982social}

Psychology
Polarity, valenced ties 
Status distinction

%Knights of the Round table, why equality conditions are fragile
%kite graph-outsider breaks mechanical solidarity
%mechanical solidarity is really an outcome of the network structure (Durkheim kind of %said that but Omar said it better). Try to get inequality in an organic solidarity system but that's unstable? (might have written that down wrong). 
%mechanical solidarity benefits from the stranger. more theoretical than empirical. 
%actually anthropologists ruined everything in the villages
%also somewhat the Marco Polo vibe, Shogun (the violation of markers of distinction, %didn't matter), The Last Samurai
%theory of subgraph cliques rather than full graph as clique
%*Tie that cuts the longest shortest path Science, 
%https://www.science.org/doi/full/10.1126/science.aau9735
%Distinction homophily, chicken or egg of

%Circles as their own, pizza pie. at least get to 0 status. Wheel can emerge
%folds vs circles
%college caves, BYU vs Notre Dame vs regulars

\input{Tabs/structfold.tex}
\input{Tabs/intraclust}
\input{Tabs/interclust}
\input{Tabs/kite}
\input{Tabs/clique}
\input{Tabs/circletriangle}
\input{Tabs/circlelong}

%Clique addition experiment

%Add cliques to star graph and see what happens

%See what happens with the caveman

\subsubsection{Long Ties and Hierarchy}
This section highlights how the hierarchy is really formed out of the redundant connections of the subgraphs. While a star graph clearly creates a hierarchy, when additional weight is added to the broker's alters, the hierarchy can be flipped, especially when they form denser subgraphs. This is in keeping with the observed structural holes above.

>>>>

\newpage
\begin{abstract}	
This paper introduces the concept of network \textit{distinction} and develops a relational metric that captures it. The metric is theoretically inspired by Bourdieu's theory of agonistic competition for distinction in social fields, where other actors and their actions provide opportunities and constraints to maximize social positioning. The basic idea behind the network metric of distinction is to subtract the \textit{autonomy} of each actor within the network from their \textit{constraint}, such that highly autonomous (and therefore less constrained) actors have more distinction. Autonomy is given by the focal actor's Eigenvector Centrality score (recording connections to well-placed others), and constraint is defined by the average Eigenvector Centrality score of each actor's neighbors in a hypothetical network where the focal actor is removed (indicating how others would benefit from that actor's exiting the network). A scalar is then applied to create mathematical and sociological equivalence between autonomy and constraint. The resulting node-level measures create a measure of \textit{distinction}, where the uniqueness of an actor's relations is more heavily rewarded than under the standard Eigenvector Centrality score. We analyze two classic datasets, Padgett and Ansell's \citeyearpar{padgett1993robust} Medici Network and Zachary's \citeyearpar{zachary1977information} Karate Club, to demonstrate that the observed actions of the agents involved are consistent with the maximization of distinction. We close by outlining how the resulting distinction measure can be used to both forecast network transformations and anticipate competitive changes resulting from external network shocks.    
\end{abstract}

@@

Bourdieu \citeyearpar{bourdieu1984distinction} most famously used the conceptualization of distinction to explain the persistence of the relationship between class and cultural hierarchies. Cultural elites in France distinguished themselves from their inferiors by adopting differentiated cultural practices that had undergone a process of consecration, which set them apart from the common. There is a strong correlation between those in the position of cultural and economic elites. Still, such a correlation is not guaranteed at the individual level, even if it is relevant at the field level.   
While largely a concept used by those who research cultural practices, the pursuit of distinction is a general expression of the interaction between status and uniqueness within a field of positions \citep{anheier1995forms}. For example, while cultural practices may be a marker of difference, the absence of association with others outside the milieu is also a matter of agency within the milieu, a question of the constraints imposed by the set of relationships within which the actor is embedded. For Bourdieu, the struggle for position in a field was a product of a field of action and a field of forces \citep{martin2003field}. The field of action refers to the agency of actors actively advancing their interests. In contrast, the field of forces describes the counteracting pressures faced by the other participants whose interests would be upset by positional transformations.

In relation to theory, network status is another name for \textit{influence} \citep{friedkin1991theoretical}, and uniqueness for \textit{autonomy} in a social environment \citep{burt1980autonomy}. Translated into network terms, Bourdieu's field of action can be understood as the status hierarchy of the network, while the field of struggle reflects the constraints that neighbors impose on an actor. An actor with high influence is well-connected to others who can be mobilized for effective action. An actor with high autonomy is not constrained by their pattern connections or forced into action by them \citep{burt76}. We often use context-specific words to denote position in the hierarchy and ability to act within that hierarchy. Nevertheless, we use ``status" and ``autonomy" below as general terms to describe the ability of actors to take action versus the level of imposition from the forces of constraint.

In developing a broader understanding and measure of social distinction, actor-level incentives and structural impediments to action become apparent. When we understand the underlying motivation that actors have to proactively reshape their network connections to achieve more distinct positions, historical patterns of network change can be better accounted for and, in some cases, even properly forecasted.

\subsection{Position as the Relational Basis for a Distinction Measure}
Bourdieu's separation of distinction into habitus and position is not a sociological problem in itself, but rather creates the analytical challenge of how an actor's network position generates dispositions and how these dispositions, in turn, create opportunities for network mobility. Both dispositions and positions are fundamentally relational outcomes in sociological terms; however, when creating a network measure of distinction, only the relational basis of position can be considered.

@@

Which tells us that the distinction of the peripheral node in the star graph is proportional to its own (scaled) status, discounted by the distinction of the star node divided by the same scaling factor.

@@

\input{Tabs/clique}
%Clique addition experiment
%Add cliques to star graph and see what happens
%See what happens with the caveman

\subsubsection{Distinction in Triadic Closure and Formation}

Triadic closure is an observed regularity in which a connection is likely to form between two actors who, although not originally connected, are first connected to a common alter (Simmel 1908). In essence, friends of friends have an increased likelihood of also becoming friends. The causes of this regularity, and the reasons for its occasional failure, are largely based on consideration of valence. Positive or negative affect drive the resulting triadic relations. The original consideration was for the cognitive balance of the individual, or the psychological congruity of affect by an individual for two objects that they see as causally connected. As such, the valenced relationship of any three actors are brought into combined consideration because any two are considered affectively combined through their joint combination as "the parts of a causal unit (Heider 1946:107)."

Beyond the cognitive congruence account of triadic closure, there is an accounting of triadic closure based on the strains new positive network connections may or may not create in existing relationships. Networks are sets of triads that are either actualized in connection, either positive or negative, or left unactuated. Actualization is largely predicated on propinquity, and the valence of theoretically possible ties never rises to the concern of a social question \citep{feld1981focused, feld1982social, mollenhorst2011shared}. However, in a contested context where ties are either positive or negative, triadic closure becomes a matter of not betraying one's friends by associating with one of their enemies, and of solidifying social alliances (Bearman 1993). However, triadic closure with a friend's friend signals commitment. The endorsement effect of triadic balance, that betrayal creates two relational problems rather than one, reinforces the surety of the actors \citep{yoon2013exchange}.

In addition to these explanations, the distinction measure also reveals the utility of a distinction based theory of triadic closure. Figure 3 \textit{Model (a)} and Table 4 present a graph and corresponding measures of near-neighbor triadic closure. The distinction of nodes 1 and 6 increases to .106, compared to 0 in the circle graph, while all other nodes fall below 0 due to the reverberating effects of the new connection. Notably, node 7 increases in eigenvector status from the combination, but overall decrease in scaled distinction due to the tie. While both eigenvector centrality and scaled distinction would predict tie formation in this simplified social structure, the scaled distinction measure would find that Node 7 would actually lose out in some ways from this arrangement. This matches with findings that initial connections do weaken when triadic closure occurs \citep{huang2018will}, consistent with a reaction to the increased constraint placed on the closure broker. This example demonstrates the fragility of the circle graph as an equality condition in a distinction-based theoretical framework, as there are opportunities for mutual increases in distinction through triadic closure. Beyond questions of balancing valence as is the current framework for understanding triadic closure, a distinction-based theoretical framework can explain the propensity towards triadic closure, but also its failure to form depending on the initial patterning of relationships.

Figure 3 \textit{Model (b)} and Table 5 present a graph and corresponding measures of a long tie formation spanning the circle graph. The connected nodes 4 and 7 both have distinction maximizing incentives to form a tie, but notably their scaled distinction increases more when they close triadically with a near neighbor. Thus, under sparse network conditions, the gains in status from local consolidation outweigh the distinction gains from increased autonomy.

%actually creates hierarchy, you only one tie to create 4 levels of hierarchy
%the pyramid is the natural form of organization
%agents form ties with maximization, then you grow transitivty 
%two conditions, inital local ties matter more, with density, long ties matter more

%Circles as their own, pizza pie. at least get to 0 status. Wheel can emerge
%folds vs circles

%kite graph-outsider breaks mechanical solidarity
%mechanical solidarity is really an outcome of the network structure (Durkheim kind of %said that but Omar said it better). Try to get inequality in an organic solidarity system but that's unstable? (might have written that down wrong). 
%mechanical solidarity benefits from the stranger. more theoretical than empirical. 
%actually anthropologists ruined everything in the villages
%also somewhat the Marco Polo vibe, Shogun (the violation of markers of distinction, %didn't matter), The Last Samurai
%theory of subgraph cliques rather than full graph as clique

%gets close to 0 negative, with non-normalized eigenvectors
%if you normalize then it all works as expected
% in the normalized kite graph, 1.0, the other get close, decreases the scalar, evs don't mean anything

%college caves, BYU vs Notre Dame vs regulars

\input{Tabs/circletriangle}
\input{Tabs/circlelong}
\input{Tabs/kite}

\input{Tabs/structfold.tex}
\input{Tabs/intraclust}
\input{Tabs/interclust}
\input{Tabs/clique}

\subsubsection{Distinction and Brokerage}

@@

\subsubsection{Long Ties and Hierarchy}

%Strogatz is random, preferential attachment of closure
%Status homophilous (enh) attachment vs preferential vs random 
%*Tie that cuts the longest shortest path Science, 
%https://www.science.org/doi/full/10.1126/science.aau9735
%Distinction homophily, chicken or egg of

This section highlights how the hierarchy is really formed out of the redundant connections of the subgraphs. While a star graph clearly creates a hierarchy, when additional weight is added to the broker's alters, the hierarchy can be flipped, especially when they form denser subgraphs. This is in keeping with the observed structural holes above.
